\documentclass[12pt, leqno]{article}
\input{common}

\begin{document}
\hdr{2023-02-22}

\section*{Orthogonal transformations and Gram-Schmidt}

We saw in the last lecture that a natural decomposition for thinking
about least squares problems is the QR decomposition
\[
  A = QR,
\]
where $Q$ is an $m \times m$ orthogonal matrix and $R$ is an $m \times
n$ upper triangular matrix.  Equivalently, we can write the
``economy'' version of the decomposition, $A = QR$ with an $m \times
n$ matrix $Q$ and an $n \times n$ upper triangular $R$, where the
columns of $Q$ form an orthonormal basis for the range space of
$A$. Using this decomposition, we can solve the least squares problem
via the triangular system
\[
  Rx = Q^T b.
\]

The {\em Gram-Schmidt} procedure is usually the first method people
learn to convert some existing basis (columns of $A$) into an
orthonormal basis (columns of $Q$).  For each column of $A$, the procedure
subtracts off any components in the direction of the previous columns,
and then scales the remainder to be unit length. In Julia, Gram-Schmidt looks
something like this:
\begin{lstlisting}
function orth_cgs0(A)
	m,n = size(A)
	Q = zeros(m,n)
	for j = 1:n
		v = A[:,j]                        # Take the jth original basis vector
		v = v-Q[:,1:j-1]*(Q[:,1:j-1]'*v)  # Orthogonalize vs q_1, ... q_j-1
		v = v/norm(v)                     # Normalize what remains
		Q[:,j] = v                        # Add result to Q basis
	end
	Q
end
\end{lstlisting}

Where does $R$ appear in this algorithm?  It appears thus:
\begin{lstlisting}
function orth_cgs(A)
	m,n = size(A)
	Q = zeros(m,n)
	R = zeros(n,n)
	for j = 1:n
		v = A[:,j]                        # Take the jth original basis vector
		R[1:j-1,j] = Q[:,1:j-1]'*v        # Project onto q_1, ..., q_j-1
		v = v-Q[:,1:j-1]*R[1:j-1,j]       # Orthogonalize vs q_1, ... q_j-1
		R[j,j] = norm(v)                  # Compute normalization constant
		v = v/R[j,j]                      # Normalize what remains
		Q[:,j] = v                        # Add result to Q basis
	end
	Q, R
end
\end{lstlisting}
That is, $R$ accumulates the multipliers that we computed from
the Gram-Schmidt procedure.  This idea that the multipliers
in an algorithm can be thought of as entries in a matrix should
be familiar, since we encountered it before when we looked at
Gaussian elimination.

\section*{Householder transformations}

The Gram-Schmidt orthogonalization procedure is not generally
recommended for numerical use.  Suppose we write $A = [a_1 \ldots
  a_m]$ and $Q = [q_1 \ldots q_m]$.  The essential problem is that if
$r_{jj} \ll \|a_j\|_2$, then cancellation can destroy the accuracy of
the computed $q_j$; and in particular, the computed $q_j$ may not be
particularly orthogonal to the previous $q_j$.  Actually, loss of
orthogonality can build up even if the diagonal elements of $R$ are
not exceptionally small.  This is Not Good, and while we have some
tricks to mitigate the problem, we need a different approach if we
want the problem to go away.

Recall that one way of expressing the Gaussian elimination algorithm
is in terms of Gauss transformations that serve to introduce zeros
into the lower triangle of a matrix.  {\em Householder} transformations
are orthogonal transformations (reflections) that can be used to similar
effect.  Reflection across the plane orthogonal to a unit normal
vector $v$ can be expressed in matrix form as
\[
  H = I-2 vv^T.
\]

Now suppose we are given a vector $x$ and we want to find a reflection
that transforms $x$ into a direction parallel to some unit vector $y$.
The right reflection is through a hyperplane that bisects the angle
between $x$ and $y$ (see Figure~\ref{fig1}), which we can construct
by taking the hyperplane normal to $x-\|x\|y$.  That is,
letting $u = x - \|x\|y$ and $v = u/\|u\|$, we have
\begin{align*}
  (I-2vv^T)x
  & = x - 2\frac{(x+\|x\|y)(x^T x + \|x\| x^T y)}{\|x\|^2 + 2 x^T y \|x\| + \|x\|^2 \|y\|^2} \\
  & = x - (x-\|x\|y) \\
  & = \|x\|y.
\end{align*}
If we use $y = \pm e_1$, we can get a reflection that zeros out all but the
first element of the vector $x$.  So with appropriate choices of reflections,
we can take a matrix $A$ and zero out all of the subdiagonal elements
of the first column.

\begin{figure}
\begin{center}
  \begin{tikzpicture}[scale=3]
    \draw[dashed] (-0.5,0) -- (1.5,0);
    \draw[thick,->] (0,0) -- (1,0.5) node [above left] {$x$};
    \draw[thick,->] (0,0) -- (1,-0.5) node [below left] {$\|x\| y$};
    \draw[thick,->] (1,-0.48) -- (1,0.48) node [right] {$x-\|x\|y$};
    \draw (0.9,0) -- (0.9,0.1) -- (1,0.1);
  \end{tikzpicture}
\end{center}
\caption{Construction of a reflector to transform $x$ into $\|x\|y$,
         $\|y\| = 1$.}
\label{fig1}
\end{figure}

Now think about applying a sequence of Householder transformations to
introduce subdiagonal zeros into $A$, just as we used a sequence of Gauss
transformations to introduce subdiagonal zeros in Gaussian elimination.
%This leads us to the following algorithm to compute the $QR$
%decomposition:
%\lstinputlisting{code/hqr1.m}
%Note that there are two valid choices of $u_1$ at each step;
%we make the choice that avoids cancellation in the obvious version
%of the formula.
%
As with $LU$ factorization, we can re-use the storage of $A$ by recognizing
that the number of nontrivial parameters in the vector $w$ at each step
is the same as the number of zeros produced by that transformation.
This gives us the following:
\begin{lstlisting}
function hqr!(A)
	m,n = size(A)
	τ = zeros(n)

	for j = 1:n

		# Find H = I-τ*w*w' to zero out A[j+1:end,j]
		normx = norm(A[j:end,j])
		s     = -sign(A[j,j])
		u1    = A[j,j] - s*normx
		w     = A[j:end,j]/u1
		w[1]  = 1.0
		A[j+1:end,j] = w[2:end]   # Save trailing part of w
		A[j,j] = s*normx          # Diagonal element of R
		tau[j] = -s*u1/normx      # Save scaling factor

		# Update trailing submatrix by multipling by H
		A[j:end,j+1:end] -= tau[j]*w*(w'*A[j:end,j+1:end])

	end

	A, tau
end
\end{lstlisting}

If we ever need $Q$ or $Q^T$ explicitly, we can always form it from
the compressed representation.  We can also multiply by $Q$ and $Q^T$
implicitly:
\begin{lstlisting}
function applyQ!(QR, τ, X)
	m, n = size(QR)
	for j = n:-1:1
		w = [1.0; QR[j+1:end,j]]
		X[j:end,:] -= tau[j]*w*(w'*X[j:end,:])
	end
	X
end

function applyQT!(QR, τ, X)
	m, n = size(QR)
	for j = 1:n
		w = [1.0; QR[j+1:end,j]]
		X[j:end,:] -= tau[j]*w*(w'*X[j:end,:])
	end
	X
end

applyQ(QR, tau, X) = applyQ!(QR, tau, copy(X))
applyQT(QR, tau, X) = applyQ(QR, tau, copy(X))
\end{lstlisting}

\section*{Givens rotations}

Householder reflections are one of the standard orthogonal
transformations used in numerical linear algebra.  The other standard
orthogonal transformation is a {\em Givens rotation}:
\[
  G = \begin{bmatrix}
    c & -s \\
    s & c
  \end{bmatrix}.
\]
where $c^2 + s^2 = 1$.  Note that
\[
  G = \begin{bmatrix}
    c & -s \\
    s & c
  \end{bmatrix}
  \begin{bmatrix}
    x \\ y
  \end{bmatrix} =
  \begin{bmatrix}
    cx - sy \\
    sx + cy
  \end{bmatrix}
\]
so if we choose
\begin{align*}
  s &= \frac{-y}{\sqrt{x^2 + y^2}}, &
  c &= \frac{x}{\sqrt{x^2+y^2}}
\end{align*}
then the Givens rotation introduces a zero in the second column.
More generally, we can transform a vector in $\bbR^m$ into a vector
parallel to $e_1$ by a sequence of $m-1$ Givens rotations, where
the first rotation moves the last element to zero, the second rotation
moves the second-to-last element to zero, and so forth.

For some applications, introducing zeros one by one is very
attractive.  In some places, you may see this phrased as a contrast
between algorithms based on Householder reflections and those based on
Givens rotations, but this is not quite right.  Small Householder
reflections can be used to introduce one zero at a time, too.
Still, in the general usage, Givens rotations seem to be the more
popular choice for this sort of local introduction of zeros.

\end{document}
